%%
\documentclass[manuscript,screen,review, 12pt]{acmart}
\let\Bbbk\relax % Fix for amssymb clash 
\usepackage{vmlmacros}
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}
\usepackage{outlines}
\begin{document}
\title{A Replacement for Pattern Matching, Inspired by Verse}

\author{Roger Burtonpatel}
\email{roger.burtonpatel@tufts.edu}
\affiliation{%
  \institution{Tufts University}
  \streetaddress{419 Boston Ave}
  \city{Medford}
  \state{Massachusetts}
  \country{USA}
  \postcode{02155}
}

\author{Norman Ramsey}
\email{nr@cs.tufts.edu}
\affiliation{%
\institution{Tufts University}
\streetaddress{177 College Ave}
\city{Medford}
\state{Massachusetts}
\country{USA}
\postcode{02155}
}

\author{Milod Kazerounian}
\email{milod.mazerouniantufts.edu}
\affiliation{%
\institution{Tufts University}
\streetaddress{177 College Ave}
\city{Medford}
\state{Massachusetts}
\country{USA}
\postcode{02155}
}

\renewcommand{\shortauthors}{Burtonpatel et al.}

\begin{abstract}
    \it{WILL BE REVISED.}
  Pattern matching is nice and has an appealing cost model, but people keep
  extending it to make it more expressive. Verse [cite-Verse] looks very simple
  and is surprisingly expressive, but the cost model is a challenge. Why not
  look for a compromise? We introduce two small functional langauges that
  attempt to address the limitations of pattern matching in popular functional
  languages.
  \end{abstract}

\maketitle

\section{Introduction}

    Pattern matching dominates observer functions. Consider the following
    implementation of \tt{List.len} in Standard ML, using observers:
    
    
    Figure 1 illustrates an example in which pattern matching is an elegant solution
    to a problem (compare to equivalent Standard ML code that manually deconstructs
    a list:)
    
    
    Most functional programmers likely prefer the first example 
    
    
    Pattern matching is a well-established and researched construct in functional
    programming langauges [cite, appeal to authority]. It shines in its ability to
    break down constructed data implicity, which is often much preferred to using
    explict deconstructors like Scheme's \tt{car} and \tt{cdr}. But when deciding to
    perform a terminal computation based on prior checks that don't involve a single
    value matching on a pattern, pattern matching struggles to retain its brevity.
    Consider the following algorithm on an abstract type~\tt{FiniteMap}, involving a
    lookup function. This examples is a modificaiton of one borrowed from SPJ
    proposal [cite]. 
    
    
    \smllst
    
    \begin{lstlisting}
    fun len yes =
        if null ys 
        then 0 
        else let xs = tl ys 
             in len xs 
             end 
    \end{lstlisting}
    \it{An implementation of List.len in Standard ML that does not use pattern 
        matching.}
    
    Manual checking and deconstruction, with built-in functions like \tt{null} and
    \tt{tl}, can be both error-prone (\bf{say more on this?}) and verbose. Most
    functional programmers likely prefer the follwing implementation of \tt{len}:
    
    \begin{lstlisting}
        fun len ys =
            case ys 
              of [] => 0
               | _::xs => 1 + len xs
        \end{lstlisting}
    
    \it{An equivalent implementation that uses pattern matching.}
    
    In most functional languages, the \tt{case} can be removed entirely in a form of
    syntactic sugar known as a \it{clausal definition}:
    \begin{lstlisting}
        fun len []      = 0
          | len (_::xs) = 1 + len xs
        \end{lstlisting}
    
    \it{A final, concise implementation that uses pattern matching.}

        \begin{verbatim}
            clunky env varl var2 | ok1 && 0k2 = val1 + val2 
                                 | otherwise = var1 + var2 
            where m1 = lookup env var1 
                  m2 = lookup env var2
                  ok1 = isJust m1 
                  ok2 = isJust m2 
                  Just val1 = m1 
                  Just val2 = m2    
        \end{verbatim}
        
        \begin{enumerate}
            \item Given two ints, look the first up in an environment which maps \tt{Int -> Maybe Int}.
            \item If the result of the first lookup is a \tt{Maybe Int val1}, look up the second name. 
            \item If the result of the second lookup is a \tt{Maybe Int val2}, return the sum of \tt{val1} and \tt{val2}. 
            \item If any step fails, return the sum of the inputs. 
        \end{enumerate}
        
        The authors show how explicit pattern matching with \tt{case} may mitigate the
        verbosity of this problem:
        
        \begin{verbatim}
            clunky env var1 var2 = 
              case lookup env var1 of 
                Nothing -> fail 
                Just val1 -> 
                  case lookup env var2 of 
                    Nothing -> fail 
                    Just val2 -> val1 + val2
            where 
                fail = var1 + var2
        \end{verbatim}
        
        They say: 
        
        "This is a bit shorter, but hardly better. Worse, if this was just one equation
        of clunky, with others that follow, then the thing would not work at all. " 
        
        The authors' suggested mitigation is pattern gaurds, which lead to the following
        solution: 
        \begin{center}
        \begin{verbatim}
            clunky env var2 var2    
            | Just vall <- lookup env var1
            , Just val2 <- lookup env var2
            = val1 + val2
        \end{verbatim}
        \it{... other equations for clunky}
        \end{center}    
        
        \it{This is obviously better. But now consider something where you need 
        or-patterns; this fails. }

        Or-patterns elminate redundant right-hand sides in a language like
        OCaml: 

    \begin{center}
    \begin{verbatim}
        let parent_game_of_token = function 
        | BattlePass      -> "Fortnite"
        | ChugJug         -> "Fortnite"
        | TomatoTown      -> "Fortnite"
        | HuntersMark     -> "Bloodborne"
        | SawCleaver      -> "Bloodborne"
        | MoghLordOfBlood -> "Elden Ring"
        | PreatorRykard   -> "Elden Ring"
        | _               -> "Irrelevant"
    \end{verbatim}
    \end{center}    

            becomes 

    \begin{center}
    \begin{verbatim}
        let parent_game_of_token = function 
        | BattlePass | ChugJug | TomatoTown  -> "Fortnite"
        | HuntersMark | SawCleaver           -> "Bloodborne"
        | MoghLordOfBlood | PreatorRykard    -> "Elden Ring"
        | _                                  -> "Irrelevant"
    \end{verbatim}
    \end{center}    

    Besides the natural appeal of brevity, or-patterns localize complexity to a
    single point. If, rather than a string, \tt{parent\_game\_of\_token}
    returned the result of a complex computation, the duplicate right-hand sides
    over multiple patterns as seen in the first model would require the
    programmer to maintain multiple redundant copies of that computation.
    \rab{should this just be the example?} Or-patterns allow programmers to
    localize such maintenance to a single point of truth. 
    \rab{todo: revise}

\verselst
    \begin{lstlisting}
        $\exists$ len. len = \ys. 
          one {ys = $\langle \rangle$; 0
               |  $\exists$ x xs. ys = $\langle$x, xs$\rangle$; 1 + len xs}
    \end{lstlisting}


    I would say 1-2 short examples that illustrate necessity of backtracking 
    and multiple results. Also, implementation of standard functions (zip): 
    maybe hard to predict runtime of functions written like this. Come up with 
    a good example. 

    We outline 2 solutions for this! 

\section{Simplification for Study and Synthesis}
\begin{outline}[enumerate]
    \1 To study "all" the interesting/standard extensions to PM, we introduce
    \PPlus.
    \2 Definition of \PPlus
    \2 {\PPlus} packages common/standard extensions to PM
    \begin{itemize}
        \item side conditions
        \item or-patterns 
        \item pattern guards 
    \end{itemize}
        \3 Examples of \PPlus. 
    \2 \PPlus\ admits of strange looking patterns, but these should not be alarming. 
    \3 \tt{when} within vcons 
    \2 -> Because they reduce to standard things by (direct) application of algebraic laws. 
        \3 K (when e) p2 ... === K \_ p2 ..., when e
        \3 when e $\leftarrow$ e === \_ <- e, when e
        \3 Laws: float out vcon, float out from guard
        \3 \bf{This law is applied repeatedly until we reach a fixed point.}
    \1 We want to study decision-making inspired by Verse. 
        \2 We remove obvious backtracking/multiple results (ask what the notecard says)
        SAY WHAT CHOICE DOES. 
        \bf{Choice forces "name knowing." Names must be known in all branches 
        if bound to unknown variable or bound to known variable if any name 
        is unknown}.
        \2 We add a classic decision-making construct: Guarded commands. 
        [cite- dijkstra paper]
        \2 The result: \VMinus. 
    \1 Definition of \VMinus
    \1 Continued Discussion:
    \2 {\VMinus} admits of many of the same pleasing computations as full Verse. 
    \2 Examples of programming in \VMinus 
    \3 Show: beautiful examples. 
    \2 {\VMinus} is at least as good as {\PPlus}? Or just as comparable? 
        Or incomparable? (Why not both?)
    \3 \bf{This is likely motivated strongly by examples of comparable code.}
    \end{outline}

IN EITHER OF THESE SECTIONS: Semantics of \PPlus\ , \VMinus

\section{(PAYLOAD) {\VMinus} can be compiled to \D 
        [Evidence that \dots efficiency]}
\begin{outline}[enumerate]
    \1 Definition of \D 
    \1 Examples of Decision Trees - From Maranget. 
    \2 Indeed, consistent with Maranget 
    \1 D is a simple generalization of Maranget, et al. 
    \1 Semantics of \D 
    \1 \D\ has a great cost model: 

    \bf{A decision tree can be exponential in size but never examines a word of
    the scrutinee more than once. }

    \1 Algorithm: Translation from \VMinus\ to \D\ : match compiler 
    \1 Theorem: Translation from \VMinus\ to \D\ preserves semantics 
    \2 Likely inductivey hypothesis. 1-4 sentences on proof max. 
\end{outline}

\section{\PPlus -> \VMinus is interesting (optional)}
\begin{outline}[enumerate]
    \1 \VMinus subsumes pattern matching 
    \1 Algorithm: Translation from \PPlus\ to \VMinus
    \1 Theorem: Translation from \PPlus\ to \VMinus preserves semantics 
    \1 Claim: Translation \PPlus\ -> \D\ is consistent with Maranget and others
\end{outline}

\section{Related Work}
\section{Future Work}
\begin{outline}[enumerate]
    \1 Why we wish for $\alpha$s
    \1 What's up with \PPlus\ ? It's worth study in its own right- in future work.
    \2 \PPlus is interesting because it combines or-patterns with pattern 
        guards. No major functional language does this. 
    \1 Programs written in Verse using ideas from \VMinus\ might have a 
    friendlier cost model (depending on compiler)
    \1 \VMinus might give Verse programmers good ideas 
        (That is, how to solve problems in Verse)
    \2 Examples of programming in Verse in the style of \VMinus 
\end{outline}

\section{Conclusion, Discussion}

\renewcommand\thesection{\Alph{section}}
\setcounter{section}{0}
\section{Proofs}
\begin{outline}
    \1 Proof: Translation from \VMinus to \D\ preserves semantics 
    \1 Proof: Translation from \PPlus\ to \VMinus preserves semantics 
    \1 Proof: Translation from \VMinus to Verse preserves semantics     
\end{outline}

\section{Trust me on \VMinus (It's reduced to Verse)}
\begin{outline}
    \1 \VMinus\ has something to do with Verse, semantically 
    \1 Our semantics of Verse is consistent with ICFP's semantics of Verse 
    \1 Definition: Our semantics of Verse
    \1 Theorem: Translation from \VMinus to Verse preserves semantics     
\end{outline}

\begin{itemize}
    \item Programmers use pattern matching. But pattern matching is not good for
    everything. 
    \item Consider Example. 
    \item Explain example. 
    \item Attempted mitigations in the past: extensions, SPJ example. 
    \item We go futher. 
    \item \bf{P3}: 
    \item Contribution: 2 languages that attempt to mitigate this problem. 
    \item Our goal: One simple, expressive, efficient decision-making construct.
    \item One is {\PPlus}, the other is {\VMinus}. 
    \item Example of {\VMinus} in action on prior example. 
\end{itemize}

We begin by informally introducing two small functional programming languages to
address the issue of inflexibility of patterns.

{\PPlus} models standard pattern matching with common extensions. {\VMinus}
models Verse without features that lead to unpredictable costs (backtracking and
multiple results) [cite-Verse]. 

To aid proofs of efficiency, we also introduce a third langauge, {\D}. {\D} is a
language of decision trees to which both {\PPlus} and {\VMinus} can be
translated. Targets of translation are efficient in the standard technical
sense: no value is ever scrutinized more than once (Maranget 2008).

To facilitate comparisons and proofs, the languages are each a subset of a
single unifying language(\cref{fig:U}). Asterisks (${}^{*}$) indicate components
of this unifying language which belong to the sub-languages. 

The keen reader will notice that our syntax includes a forms for decision trees,
which we have not yet discussed. Decision trees will be crucial in proving
properties of efficiency of our languages, and will we be discussed in full in a
later section. 

\begin{figure}[h!p]
    \small
    \begin{flushleft}
        \begin{bnf}
        $P$ : \textsf{Programs} ::=
        $\bracketed{d}$ : definition
        ;;
        $d$ : \textsf{Definitions} ::=
        | $\mathit{val}\; x\; \mathit{=}\; \expr$ : bind name to expression
        ;;
        $\expr$ : Expressions ::= 
        | $x, y, z$             : names
        | $K\bracketed{\expr}$  : value constructor application 
        | $\lambda x.\; \expr$  : lambda declaration  
        | $\expr[1]\; \expr[2]$ : function application 
    
        | $\mathit{case}\; \expr\; \bracketed{p\; \rightarrow\; \expr}$ : $\rm{case expression}^{*}$
        | $\mathit{if}\; \mathit{[}\; g\; 
            \bracketed{[] g}\; \mathit{]}\; \mathit{fi}$                : $\rm{if-fi}^{**}$
        | $\dt$                                                         : $\rm{decision tree}^{***}$
        ;;
        $\v$ : Values ::= 
          $K\bracketed{\v}$     : value constructor application 
        | $\lambda x.\; \expr$  : lambda value 
        ;;
        $p$ : $\textsf{Patterns}^{*}$ ::= 
        $x$ : name 
        | $K\; \bracketed{p}$           : value constructor application 
        | $\mathit{when}\; \expr$       : side condition
        | $p, p'$                       : pattern guard 
        | $p\; \leftarrow\; \expr$      : pattern from explicit expression  
        | $p_{1}\pbar p_{2}$            : or-pattern
        ;;
        $g$ : $\textsf{Guarded Expressions}^{**}$ ::=  
        $\boldsymbol{\rightarrow}\expr$ : terminating experession
        | $\expr;\; g$                  : intermediate expression 
        | $\vexists{x} g$      : existential 
        | $x = \expr;\; g$              : equation 
        ;;
        \dt : $\textsf{Decision Trees}^{***}$ ::= 
        | $\mathit{case}\; x\; \mathit{of}\; 
            \bracketed{\vert\; K\bracketed{x}\; \mathit{=>}\; \dt} 
            [\vert\; x\; \mathit{=>} \dt]$                              : $\rm{test node }^{***}$
        | $\expr$                                                       : $\rm{match node }^{***}$
        | $\mathit{if}\; x\; \mathit{then}\; \dt\; \mathit{else}\; \dt$ : $\rm{condition with two children }^{***}$
        | $\mathit{let}\; x\; \mathit{=}\; \expr\; \mathit{in}\; \dt$   : $\rm{let-bind a name}^{***}$
        % ;;
        % $K$ : \textsf{Value Constructors} ::=
        % | $\mathit{true}\; \vert\; \mathit{false}$ : booleans
        % | $\mathit{\#}x$                           : name beginning with $\mathit{\#}$
        % | $\mathit{A-Z}x$                          : name beggining with capital letter
        % | $[\mathit{-}\vert\mathit{+}]
        %     (\mathit{0}-\mathit{9})+$              : signed integer literal 
        \end{bnf}
        \medskip
        
        
        \it{Concrete Syntax}: "$\lambda$" and "$\exists$" each scope as far to
        the right as possible.
        
        $\hskip 8em$ For example, $(\lambda y.\> \vexists{x}\> x = 1;\> x + y)$ means 
        $(\lambda y.\> (\vexists{x}\> ((x = 1);\> (x + y))))$.
        
        Parentheses may be used freely to aid readability and override default precedence.

        A \it{name} is any token that is not an integer literal, does not
        contain whitespace, a bracket, or parenthesis, and is not a value
        constructor name or a reserved word.
        
        \medskip

        % \bf{Desugaring of Extended Expressions}

        ${}^{*}$ Indicates forms within {\PPlus}

        ${}^{**}$ Indicates forms within {\VMinus}
        
        ${}^{***}$ Indicates forms within {\D}

    \end{flushleft}
    
    \medskip

    

    \caption{\U, a decision-making language}
    \Description{A BNF grammar for \U, the universal decision-making language.
                 It includes patterns, guarded expressions, and decision trees.}
    \label{fig:U}
\end{figure}
\begin{table}[h]
    \centering
    \small
    \begin{tabular}{l l l l}
        \textbf{Syntactic Forms} & \textbf{Cases} & \textbf{Belong to} \\

        $P$ : Programs & $\bracketed{d}$ & many definitions & \\
        $d$ : Definitions & $\mathit{val}\; x\; \mathit{=}\; \expr$ & bind name to expression & \\
        $\expr$ : Expressions & $x, y, z$ & names & \\
        & $K\bracketed{\expr}$ & value constructor applied to expressions & \\
        & $\lambda x.\; \expr$ & lambda declaration & \\
        & $\expr[1]\; \expr[2]$ & function application & \\
        & $\mathit{case}\; \expr\; \bracketed{p\; \rightarrow\; \expr}$ & $\rm{case expression}$ & \\
        & $\mathit{if}\; \mathit{[}\; g\; \bracketed{[] g}\; \mathit{]}\; \mathit{fi}$ & $\rm{if-fi}$ & \\
        & $\dt$ & $\rm{decision tree}$ & \\
        $\v$ : Values & $K\bracketed{\v}$ & value constructor applied to values & \\
        & $\lambda x.\; \expr$ & lambda value & \\
        $p$ : Patterns & $x$ & name & \\
        & $K\; \bracketed{p}$ & value constructor applied to patterns & \\
        & $\mathit{when}\; \expr$ & side condition & \\
        & $p, p'$ & pattern guard & \\
        & $p\; \leftarrow\; \expr$ & pattern from explicit expression & \\
        & $p_{1}\pbar p_{2}$ & or-pattern & \\
        $g$ : Guarded Expressions & $\boldsymbol{\rightarrow}\expr$ & terminating expression & \\
        & $\expr;\; g$ & intermediate expression & \\
        & $\vexists{x} g$ & existential & \\
        & $x = \expr;\; g$ & equation & \\
        \dt : Decision Trees & $\mathit{case}\; x\; \mathit{of}\; \bracketed{\vert\; K\bracketed{x}\; \mathit{=>}\; \dt} [\vert\; x\; \mathit{=>} \dt]$ & $\rm{test node}$ & \\
        & $\expr$ & $\rm{match node}$ & \\
        & $\mathit{if}\; x\; \mathit{then}\; \dt\; \mathit{else}\; \dt$ & $\rm{condition with two children}$ & \\
        & $\mathit{let}\; x\; \mathit{=}\; \expr\; \mathit{in}\; \dt$ & $\rm{let-bind a name}$ & \\
    \end{tabular}
    \caption{Example LaTeX Table}
    \label{tab:example}
\end{table}

\begin{table}[h]
    \centering
    \small
    \begin{tabular}{l l l}
        \textbf{Syntactic Forms} & \textbf{Cases} & \textbf{Unique to} \\
        \hline
        $P$ : Programs & $\bracketed{d}$ & \\
        $d$ : Definitions & $\mathit{val}\; x\; \mathit{=}\; \expr$ & \\
        $\expr$ : Expressions & $x, y, z$ & \\
        & $K\bracketed{\expr}$ &  \\
        & $\lambda x.\; \expr$ & \\
        & $\expr[1]\; \expr[2]$ & \\
        & $\mathit{case}\; \expr\; \bracketed{p\; \rightarrow\; \expr}$ & \PPlus \\
        & $\mathit{if}\; \mathit{[}\; g\; \bracketed{[] g}\; \mathit{]}\; \mathit{fi}$ & \VMinus \\
        & $\dt$ & \D \\
        $\v$ : Values & $K\bracketed{\v}$ & \\
        & $\lambda x.\; \expr$ & \\
        $p$ : Patterns & $x$ & \PPlus \\
        & $K\; \bracketed{p}$ & \PPlus \\
        & $\mathit{when}\; \expr$ & \PPlus \\
        & $p, p'$ & \PPlus \\
        & $p\; \boldsymbol{\leftarrow}\; \expr$ & \PPlus \\
        & $p_{1}\pbar p_{2}$ & \PPlus \\
        $g$ : Guarded Expressions & $\boldsymbol{\rightarrow}\expr$ & \VMinus \\
        & $\expr;\; g$ & \VMinus \\
        & $\vexists{x} g$ & \VMinus \\
        & $x = \expr;\; g$ & \VMinus \\
        \dt : Decision Trees & $\mathit{case}\; x\; \mathit{of}\; \bracketed{\vert\; K\bracketed{x}\; \mathit{=>}\; \dt} [\vert\; x\; \mathit{=>} \dt]$ & \D \\
        & $\expr$ & \D \\
        & $\mathit{if}\; x\; \mathit{then}\; \dt\; \mathit{else}\; \dt$ & \D \\
        & $\mathit{let}\; x\; \mathit{=}\; \expr\; \mathit{in}\; \dt$ & \D \\
    \end{tabular}
    \caption{Example LaTeX Table}
    \label{tab:example}
\end{table}



Most syntactic categories in \U are present in all of its subsets: Values, Value
Constructors, Definition Forms, and most forms of expression are shared. Indeed,
the sub-langauge, like in Verse, is just the the lambda calculus with a few
extended syntactic categories (this time favoring value constructors over
tuples). Like in Verse, every Lambda Calculus program is a valid \U program. 

On top of this core, three languages are defined as subsets of \U. {\PPlus} is
the subset of~\U including Patterns and the $\mathit{case}\; \expr\;
\bracketed{p \rightarrow \expr}$ form of case expression. {\VMinus} is the
subset of \U including Guarded Expressions and \it{if-fi}. Finally, {\D} is the
subset of \U that includes decision tree syntax, including the reduced case
expression, $\mathit{case}\; x\; \mathit{of}\; \bracketed{\vert\;
K\bracketed{x}\; \mathit{=>}\; \dt} [\vert\; x\; \mathit{=>} \dt]$.
Importantly, the three are mutually exclusive with respect to exactly these
extensions; all subsets share the sub-langauge and no subset has more than one
of the expanded \it{case}, \it{if-fi}, or the decision tree syntactic
categories. 

% {\PPlus} provides or-patterns, side-conditions, and pattern guards, whose
% combination does not appear in Haskell, OCaml, Standard ML, or any other major
% functional language. 
% \bigskip
% Next steps: Finish introduction of language table, say where semantics are. 


\it{The following sections and paragraphs are written:}

\begin{itemize}
    \item \bf{Subsection: General Evaluation}
    \item Big-step opsem with environments. 
    \item \bf{Subsection: {\PPlus}}
    \item Patterns, case, environment + disjoint union 
\end{itemize}

\it{The following sections and paragraphs are in progress:}
\begin{itemize}
    \item when, or-patterns, pattern guards 
    \item \bf{Subsection: {\VMinus}}
    \item Guarded expressions, new type of environment
    \item Evaluation stragety 
\end{itemize}

\it{Finally, much has been commented out below because I am still determining
    the order in which their appearance is most clear. After your review of this
    initial format, I will begin to include them.}


Prior work has introduced extesions to pattern matching [cite, including
SPJ proposal for pattern guards.]

{\PPlus} provides or-patterns, side-conditions, and pattern guards, whose
combination does not appear in Haskell, OCaml, Standard ML, or any other major
functional language. 

\bigskip
interesting thoughts: 

Having a strategem for verse to decision tree is analagous to tail recursion. 
You have to write your code in a way that allows the optimization. 


Pattern matching is a well-established paradigm within functional languages, and
"""has been the subject of significant discourse.""" \it{Appeal to authority here.}

Without pattern matching, it can be tiresome to deconstruct data using manual 
accessor functions. Consider the following Standard ML code: 
\smllst

\begin{lstlisting}
val rec len = fn ys => 
    if null ys 
    then 0 
    else let xs = tl ys 
         in len xs 
         end 
\end{lstlisting}
\it{An implementation of List.len in Standard ML that does not use pattern 
    matching.}

Manual checking and deconstruction, with built-in functions like \tt{null} and
\tt{tl}, can be both error-prone (\bf{say more on this?}) and verbose. Most
functional programmers likely prefer the follwing implementation of \tt{len}:

\begin{lstlisting}
    val rec len = fn ys => 
        case ys 
          of [] => 0
           | _::xs => 1 + len xs
    \end{lstlisting}

\it{An equivalent implementation that uses pattern matching.}

Indeed, pattern matching is quite appealing when deconstructing data, especially 
data that may be represented with an algebraic data type, is the primary programming problem: 



\begin{lstlisting}
val len = \ys. case ys 
                 of [] -> 0
                  | _::xs -> 1 + len xs
\end{lstlisting}


Figure 1 illustrates an example in which pattern matching is an elegant solution
to a problem (compare to equivalent Standard ML code that manually deconstructs
a list:)


Most functional programmers likely prefer the first example 

However, language designers continue to extend pattern matching 

\section{Pattern Matching as it is Now}
\it{Pattern matching} is defined as "checking and locating of specific sequences
of data of some pattern among raw data or a sequence of tokens." We will return
to the notion of "checking" often in this paper: pattern matching answers the
question "when I'm checking to see if a piece of data (called a \it{scrutinee})
is of the same form as a certain pattern, does that match succeed or fail?"

\it{Example}.

In addition to this checking and locating, pattern matching serves as
\it{assignment}: it can bind fresh variables based on the form of data and use
those bindings in subsequent expressions. 

\it{Example}.

Here, "checking" means "do the data match what I expect them to." Because
pattern matching is inherently built to match a scrutinee (pure data) with a
pattern, pattern matching is quite expressive in these cases. 

Because of this, most modern functional languages, especially \it{data
dependency languages} like Haskell or ML, \it{(is this right?)} employ pattern
matching as their main way to deconstruct data (citation?). 

We here explore pattern matching through the lens of {\PPlus}, an invented
language that has pattern matching along with several of its popular extensions.
(The examples above are written in {\PPlus}.) 
\subsection{Strengths}

- "Checking" and assignment- nice! No car, cdr 

- Nested patterns - powerful 

- Literal patterns let you mix names and values 

\subsection{Weaknesses and Proposed Mitigations}

We return to the concept of unifying "checking" with assignment, i.e. "match if
the data take this form, and give them names." Pattern matching succeeds here
when checking means "is the form of data the way I expect"; in fact, as we know,
assignment in general \it{is} pattern matching (figure/example?). But when
checking means "does this computation succeed" or "does this binding conflict
with a prior binding," pattern matching is at a loss, where Verse succeeds. 

Example: 


Pattern matching's extensions get closer to unifying "checking" and binding.
Here, Verse enjoys a different suite of advantages. 

First, its "or" operation (`one` with `choice`) allows for more than patterns to
appear as a top-level "choose this or that" construct in a match sequence; you
can also include arbitrary expressions. You can't do this in pattern matching's
version, which is an or-pattern. Simply put, saying "does this pattern match or
is this expression true" is easy in Verse and clunky if you use patterns. I'll
show examples at our meeting. 

Second, Verse can express operations out of order, letting important checks
appear higher up even if they are executed later. This helps program legibility.
Again, I have examples from the chapter I wrote today. 

Third, in a pattern-match clause, the initial data must still match an initial
pattern in order to enter a guard; in Verse there is no restriction. This is
minor, because you could simply match the data to a variable, and then enter a
guard-- but again, all of these advantages are in elegance and brevity, and
elegant that solution is not. 

Finally, mingling pattern guards with other extensions to pattern matching
(especially or-patterns) is a murky subject. Haskell has pattern guards and side
conditions, but no or-patterns. OCaml has side conditions and or-patterns, but
no guards. Mixing all three is (according to some readings) simply difficult for
implementers- including, interestingly, those of parsers. In Verse, having `one`
and `choice` closely tied in with the simple `e1 = e2` equation form, which by
itself subsumes pattern matching, side-conditions, and pattern guards, means
that integrating options is easy. A key theme that arises of this: Verse has
fewer constructs, and they are more expressive.  

\section{A Proposal, Inspired by the Verse Calculus}

\subsection{Verse Flexibility}
\subsection{Something else}
\subsection{A third thing}

\section{Verse's Equations Subsume Pattern Matching}

\subsection{Claim}
\subsection{Proof}
\subsection{Translations}

\section{(Maybe) Writing Efficient Verse Code}

\subsection{Claim}
\subsection{Proof}
\subsection{Translations}


\section{Citations and Bibliographies}

\end{document}
\endinput
%%
