\documentclass[manuscript,screen,review, 12pt, nonacm]{acmart}
\let\Bbbk\relax % Fix for amssymb clash 
\usepackage{vmlmacros}
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}
\usepackage{outlines}
\setlength{\headheight}{14.0pt}
\setlength{\footskip}{13.3pt}
\title{An Alternative to Pattern Matching, Inspired by Verse}
\author{Roger Burtonpatel}
\email{roger.burtonpatel@tufts.edu}
\affiliation{%
  \institution{Tufts University}
  \streetaddress{419 Boston Ave}
  \city{Medford}
  \state{Massachusetts}
  \country{USA}
  \postcode{02155}
}
\begin{document}



\section{Pattern Matching and equations}
\label{pmandequations}

In this section I'll flush out the definitions, forms, and tradeoffs of pattern
matching and equations. This will inform the compromises I make in \VMinus\ down
the line in Section~\ref{vminus}.

\subsection{Pattern matching}
\label{pmoverobservers}

% \1 What is pattern matching 

Pattern matching lets programmers examine and deconstruct data by matching them
against patterns. When a pattern $p$ matches with a value $v$, it can produce
bindings for any sub-values of $v$. 

Why use pattern matching? What could programmers use instead? One tool a
programmer might use to deconstruct data is \it{observers}[cite- Liskov and
Guttag]: functions that explicitly inquire a piece of data's structure and
manually extract its components. Examples of observers in functional programming
languages include Scheme's \tt{null?}, \tt{car}, and \tt{cdr}, and ML's
\tt{null}, \tt{hd}, and \tt{tl}. But many function programmers favor pattern
matching. I claim to know a reason why: \it{code written using pattern matching
and its extensions has superior properties to code written using observers}.
I've listed these below, and I'll refer to them as the Nice Properties from here
on out: 
\bf{\begin{enumerate}
    \item With pattern matching, code more closely resembles algebraic laws. 
    \label{p1}
    \item With pattern matching, it's easier to avoid duplicating code.
    \label{p2}
    \item Pattern matching plays nicer with user-defined types. 
    \label{p3}
    \item Pattern matching militates toward naming important intermediate
    values. \nolinebreak
    \label{p4}
    \item With pattern matching, a compiler may be able to tell if the code
    omits an important case. 
    \label{p5}
\end{enumerate}}
I'll demonstrate how pattern matching creates code that demonstrates the Nice
Properties, even in a tiny example. Consider a \it{shape} datatype in Standard
ML, which represents shapes by their side lengths: 

\medskip 
\begin{minipage}[t]{\textwidth}
    \begin{verbatim}
        datatype shape = SQUARE of real 
                       | TRIANGLE of real * real 
                       | TRAPEZOID of real * real * real
\end{verbatim}
\end{minipage}
\medskip 

Let's define an \tt{area} function on \it{shape}s, with this type and these
algebraic laws: 

\medskip 
\begin{minipage}[t]{\textwidth}
    \begin{verbatim}
        area : shape -> real 
        (* area (SQUARE s)              == s * s 
           area (TRIANGLE (w, h))       == 0.5 * w * h
           area (TRAPEZOID (b1, b2, h)) == 0.5 * b1 * b2 * h
        *)
\end{verbatim}
\end{minipage}
\medskip 

Now consider two implementations of \tt{area}, one with observers and one with
pattern matching, in Figure~\ref{fig:area}.

    \begin{figure}[H]
      \begin{minipage}[t]{0.7\textwidth}
        \begin{verbatim}
fun area sh =
  if isSquare sh
  then sqSide sh * sqSide sh
  else if isTriangle sh 
  then 0.5 * triW sh * triH sh
  else 0.5 * traB1 sh * traB2 sh * traH sh
            \end{verbatim}
            \Description{An implementation of area observers}
        \subcaption{\tt{area} with observers}
            \label{fig:observerarea} 
      \end{minipage}
      \vfill
      \begin{minipage}[t]{0.7\textwidth}
        % \centering 
        \begin{verbatim}
fun area sh =
  case sh 
    of SQUARE s              => s * s
     | TRIANGLE (w, h)       => 0.5 * w * h
     | TRAPEZOID (b1, b2, h) => 0.5 * b1 * b2 * h
                \end{verbatim}
       \Description{An implementation of area using implicit deconstruction via
       patterns.}
       \vspace{2.2em}
       \subcaption{\tt{area} with pattern matching}
       \label{fig:pmarea}
      \end{minipage}
      \caption{Implementing \tt{area} using observers is tedious, and the code
      doesn't look like the algebraic laws. Using pattern matching makes an
      equivalent implementation more appealing.}
      \label{fig:area}
    \end{figure}

    Implementing the observers \tt{isSquare}, \tt{isTriangle}, \tt{sqSide},
    \tt{triW}, \tt{traB1}, \tt{traB2}, and \tt{traH} is left as an
    (excruciating) exercise to the reader. 
    
    If that prospect doesn't convince you that pattern matching avoids a lot of
    the problems that observers have, let's see how each of our Nice Properties
    holds up in \tt{area}: 
    
    \begin{enumerate}
        \item[\bf{(1)}] \ref{fig:pmarea}, which uses pattern matching, more
        closely resembles the algebraic laws for \tt{area}. Good! 
        \item[\bf{(2)}] \ref{fig:observerarea} had to call observers like
        \tt{squareSide} multiple times, and each observer needs \tt{sh} as an
        argument. \ref{fig:pmarea} was able to extract the \tt{shape}s' internal
        values with a single pattern, and the name \tt{sh} is not duplicated
        anywhere in its body. Again, \ref{fig:pmarea} wins. 
        \item[\bf{(3)}] Where did \tt{isSquare}, \tt{sqSide}, and all the other
        observers come from? To even \it{implement} \ref{fig:observerarea}, a
        programmer has to define a whole new set of observers for \tt{shape}s!
        Most programmers find this tiresome indeed. \ref{fig:pmarea} did not
        have to do any of this. 
        \item[\bf{(4)}] To extract the internal values, \ref{fig:pmarea} had to
        name them, and their names serve as documentation. 
        \item[\bf{(5)}] If the user adds another value constructor to
        \tt{shape}- say, \tt{CIRCLE}, \ref{fig:observerarea} will not cause the
        compiler to complain, and if it's passed a \tt{CIRCLE} at runtime, the
        program will likely crash! In \ref{fig:pmarea}, the compiler will warn
        the user of the possibility of a \tt{Match} exception, and even tell
        them that they must add \tt{CIRCLE} to rule out this possibility. 
    \end{enumerate}

    Having had the chance to compare Figures \ref{fig:observerarea}
    and~\ref{fig:pmarea}, if you moderately prefer the latter, that's good: most
    functional programmers- in fact, most \it{programmers}- likely do as well. 

    Figure~\ref{fig:pmarea} provides an opportunity to introduce a few terms
    that are common in pattern matching. \ref{fig:pmarea}~is a classic example
    of where pattern matching most commonly occurs: within a case expression.
    Case expressions test a \it{scrutinee} (\tt{sh} in~\ref{fig:pmarea}) against
    a list of patterns (\tt{SQUARE s}, etc.). When the result of evaluating the
    scrutinee matches with a pattern, the program evaluates the right-hand side
    of the respective branch (\tt{s * s}, etc.) within the case
    expression.\footnote{OCaml, which you'll see in future sections, calls case
    \tt{match}. Some literature\cite{guardproposal} calls this a \it{head
    expression}. I follow Ramsey's\cite{bpc} and Maragnet's\cite{maranget}
    example in calling the things \it{case} and \it{scrutinee}. Any of these
    terms does the job.} 
% \1 Pattern matching has these properties, observers do not. 

\subsection{Popular extensions to pattern matching}
\label{extensions}

    Extensions to pattern matching simplify cases that are otherwise
    troublesome. Specifically, extensions help restore the Nice Properties in
    cases where pattern matching fails to satisfy them. 
    
    In this section, I illustrate several such instances of exactly this, and
    demonstrate how extensions help programmers write code that adheres to the
    Nice Properties. The three extensions I describe are those commonly found in
    the literature and implemented in common compilers: side conditions, pattern
    guards, and or-patterns. 
    
    For the sake of comparison, I coin the term \it{bare pattern matching} to
    denote pattern matching \it{without} extensions: in bare pattern matching,
    the only syntactic forms of patterns are names and applications of value
    constructors to zero or more patterns. 


% Subject: Extensions to pattern matching. 

\subsubsection{Side conditions}

    First, I illustrate why programmers want \it{side conditions}, an extension
    to pattern matching common in most popular functional programming languages,
    including OCaml, Erlang, Scala, and~Haskell\footnote{I use the term \it{side
    conditions} to refer to a pattern followed by a boolean expression. Some
    languages call this a \it{guard}, which I use to describe a different, more
    powerful extension to pattern matching in Section~\ref{guards}.  
    Technically, Haskell \it{only} has guards, but they subsume side conditions,
    so I hand-wavingly say that it does have side conditions.}. 
    
    Let's dive in! I'll define a (rather silly) function \tt{exclaimTall} in
    OCaml on \tt{shape}s. I'll have to translate our \tt{shape} datatype to
    OCaml, and while I'm at it, I'll write the type and algebraic laws for
    \tt{exclaimTall}:

    \begin{minipage}[t]{\textwidth}        
        \centering 
        \begin{verbatim}

type shape = Square of float 
           | Triangle of float * float
           | Trapezoid of float * float * float

exclaimTall : shape -> string 
(*
exclaimTall (Square s)              == "Wow! That's a tall square!", 
                                        where s > 100.0
exclaimTall (Triangle (w, h))       == "Goodness! Towering triangle!",
                                        where h > 100.0
exclaimTall (Trapezoid (b1, b2, h)) == "Zounds! Tremendous trapezoid!", 
                                        where h > 100.0
exclaimBigArea sh                   == "Your shape is small.", 
                                        otherwise
*)
    \end{verbatim}
    \end{minipage}

    Armed with pattern matching, I'll try to implement \tt{exclaimTall} in OCaml
    (Figure~\ref{fig:ifexclaimtall}).

    \begin{figure}[ht]
        \begin{verbatim}
            let exclaimTall sh =
            match sh with 
              | Square s -> if s > 100.0 
                            then "Wow! That's a tall square!"
                            else "Your shape is small." 
              | Triangle (_, h) -> 
                            if h > 100.0 
                            then "Goodness! Towering triangle!"
                            else "Your shape is small." 
              | Trapezoid (_, _, h) -> 
                            if h > 100.0
                            then "Zounds! Tremendous trapezoid!"
                            else "Your shape is small." 
            \end{verbatim}    
        \caption{An invented function \tt{exclaimTall} in OCaml combines pattern
        matching with an \tt{if} expression, and is not very pretty.}   
        \Description{An implementation of a function exclaimTall in OCaml, which
        uses pattern matching and an if statement.}
        \label{fig:ifexclaimtall}
    \end{figure}
    
    Here, I'm using the special variable \tt{\_}- a wildcard pattern- to
    indicate that I don't care about the bindings of a pattern. 

    I (and hopefully you, the reader) are not thrilled with this code. It gets
    the job done, but it fails to adhere to Nice Properties \ref{p1}
    and~\ref{p2}: the code does not look like the algebraic laws, and it
    duplicates right-hand side, \tt{"Your shape is small"}, three times. I find
    the code unpleasant to read, too: the actual “good” return values of the
    function, the exclamatory strings, are gummed up in the middle of the
    \tt{if-then-else} expressions.
    
    Fortunately, this code can be simplified by using the \tt{shape} patterns
    with a \it{side condition}, i.e., a syntactic form for “match a pattern
    \it{and} a boolean condition.” The \tt{when} keyword in OCaml provides such
    a form, as seen in Figure~\ref{fig:whenexclaimtall}.
        
        \begin{figure}[]
            \begin{verbatim}
                let exclaimTall sh =
                  match sh with 
                    | Square s when s > 100.0 ->
                            "Wow! That's a tall square!"
                    | Triangle (_, h) when h > 100.0 ->
                            "Goodness! Towering triangle!"
                    | Trapezoid (_, _, h) when h > 100.0 -> 
                            "Zounds! Tremendous trapezoid!"
                    | _ ->  "Your shape is small." 
                \end{verbatim}
            \caption{With a side condition, \tt{exclaimTall} in OCaml becomes
            simpler and more adherent to the Nice Properties.} 
            \Description{An implementation of a function exclaimTall in OCaml,
            which uses pattern matching and a side condition.}
            \label{fig:whenexclaimtall}
        \end{figure}

    A side condition streamlines pattern-and-boolean cases and minimize
    overhead. And a side condition can exploit bindings that emerge from the
    preceding pattern match. For instance, the \tt{when} clauses in
    Figure~\ref{fig:whenexclaimtall} exploit names \tt{s} and \tt{h}, which are
    bound in the match of \tt{sh} to \tt{Square s}, \tt{Triangle (\_, h)}, and
    \tt{Trapezoid (\_, \_, h)}, respectively. 

    A side condition can incorporate an extra “check”- in this case, a boolean
    expression- within a pattern. But side conditions have a limitation. The
    check can make a decision based off of an expression evaluating to \tt{true}
    or \tt{false}. But it can't make a decision based off an arbitrary pattern
    match, and it can't bind names for the programmer to use in the right-hand
    side. In the next section, I showcase when this limitation matters, and how
    another extension addresses it. 

    \subsubsection{Pattern guards}
    \label{guards}

    To highlight a common use of pattern guards, I modify an example from Erwig
    \& Peyton-Jones\cite{guardproposal}, the proposal for pattern guards in GHC.
    Suppose I have an OCaml abstract data type of finite maps, with a lookup
    operation: 

    \medskip
    \begin{minipage}[t]{\textwidth}
        \centering 
        \begin{verbatim}
            lookup : finitemap -> int -> int option
        \end{verbatim}
    \end{minipage}
    
    Let's say I want to perform three successive lookups, and call a “fail”
    function if either of them returns \tt{None}. I want a function with 
    this type and algebraic laws: 

    \begin{minipage}[t]{\textwidth}
        \centering 
        \begin{verbatim}
          

            tripleLookup : finitemap -> int -> int

            tripleLookup rho x == z, where 
                                       lookup rho x == Some w
                                       lookup rho w == Some y
                                       lookup rho y == Some z
            tripleLookup rho x == handleFailure x, otherwise
            
            handleFailure : int -> int 
            (* handleFailure's implementation is unimportant.
               handleFailure (x : int) = ... some error-handling ... -> x *)  

        \end{verbatim}
    \end{minipage}

    To express this computation succinctly, the program needs to make decisions
    based on how successive computations match with patterns, but neither bare
    pattern matching nor side conditions don't give that flexibility. 
    
    Side conditions don't appear to help here, so I'll try with bare pattern
    matching. Figure~\ref{fig:pmtriplelookup} shows how I might implement
    \tt{tripleLookup} as such. 

    \begin{figure}[ht]
        \begin{verbatim}

          let tripleLookup (rho : finitemap) (x : int) =
              match lookup rho x with Some w -> 
                (match lookup rho w with Some y -> 
                  (match lookup rho y with Some z -> z
                  | _ -> handleFailure x)
                | _ -> handleFailure x)
              | _ -> handleFailure x
            \end{verbatim}
        \caption{\tt{tripleLookup} in OCaml with bare pattern matching breaks
                    Nice Property~\ref{p2}: avoiding duplicated code. } 
                    
        \Description{An implementation of the tripleLookup with three levels of
        nested pattern matching.}
        \label{fig:pmtriplelookup}
    \end{figure}

    Once again, the code works, but it's lost Nice Properties \ref{p2}
    and~\ref{p1} by duplicating three calls to \tt{handleFailure} and stuffing
    the screen full of syntax that distracts from the algebraic laws.
    Unfortunately, it's not obvious how a side condition could help us here,
    because we need pattern matching to extract and name internal values from
    constructed data.

    To restore the Nice Properties, I'll introduce a more powerful extension to
    pattern matching: \it{pattern guards}, a form of “smart pattern” in which
    intermediate patterns bind to expressions within a single branch of a
    \tt{case}. Pattern guards can make \tt{tripleLookup} \it{much} appear
    simpler, as shown in Figure~\ref{fig:guardtriplelookup}-- which, since
    pattern guards aren't found in OCaml, is written in Haskell.

    \begin{figure}[hbt!]  
        \begin{center}
        \begin{verbatim}
                        tripleLookup rho x
                           | Just w <- lookup rho x
                           , Just y <- lookup rho w
                           , Just z <- lookup rho y
                           = z
                        tripleLookup _ x = handleFailure x
        \end{verbatim}
        \end{center}    
    \caption{Pattern guards swoop in to restore the Nice Properties, and all is
    right again.} 
    \Description{An implementation of nameOf using pattern guards.}
    \label{fig:guardtriplelookup}
    \end{figure}

    Guards appear as a comma-separated list between the \tt{|} and the \tt{=}.
    On the left-hand side of the \tt{<-} is a pattern, and on the right is an
    expression. At runtime, the program processes a guard by evaluating the
    expression and testing if it matches with the pattern. If it does, it
    processes the next guard with any bindings introduced by processing guards
    before it. If it fails, the program skips evaluating the rest of the branch
    and falls through to the next one. As a bonus, a guard can simply be a
    boolean expression which the program tests the same way it would a side
    condition, so guards subsume side conditions! 
    
    If you need further convincing on why programmers want for guards, look no
    further than Erwig \& Peyton-Jones's \it{Pattern Guards and Transformational
    Patterns}\cite{guardproposal}, the proposal for pattern guards in GHC: the
    authors show several other examples where guards drastically simplify
    otherwise-maddening code. 
    
    The power of pattern guards lies in the ability for expressions within
    guards to utilize names bound in preceding guards, enabling imperative
    pattern-matched steps with expressive capabilities akin to Haskell's \tt{do}
    notation. It should come as no surprise that pattern guards are built in to
    GHC. 

\subsubsection{Or-patterns}

    I conclude our tour of extensions to pattern matching with or-patterns,
    which are built in to OCaml. Let's consider a final example. I have a type
    \tt{token} which represents a token in a video game and how much fun it is,
    and need to quickly know what game it's from and how fun I'd have playing
    it. To do so, I'm going to write a function \tt{game\_of\_token} in OCaml.
    Here are the \tt{token} type and the type and algebraic laws for
    \tt{game\_of\_token}. 

\begin{minipage}[t]{\textwidth}
    \centering 
    \begin{verbatim}

    type funlevel = int

    type token = BattlePass  of funlevel 
               | ChugJug     of funlevel 
               | TomatoTown  of funlevel
               | HuntersMark of funlevel 
               | SawCleaver  of funlevel
               | MoghLordOfBlood of funlevel 
               | PreatorRykard   of funlevel
               ... other tokens ...
                   

game_of_token : token -> string * funlevel

game_of_token t == ("Fortnite", f),  where t is any of 
                                      BattlePass f, 
                                      ChugJug f, or
                                      TomatoTown f
game_of_token t == ("Bloodborne", 2 * f), 
                                      where t is any of 
                                        HuntersMark f or 
                                        SawCleaver f
game_of_token t == ("Elden Ring", 3 * f), 
                                      where t is any of 
                                        MoghLordOfBlood f or  
                                        PreatorRykard f
game_of_token _ == ("Irrelevant", 0), otherwise
    \end{verbatim}
\end{minipage}        
        
        I can write code for \tt{game\_of\_token} in OCaml using bare patterns
        (Figure~\ref{fig:baregot}), but I'm dissatisfied with how it fails the
        first three (\ref{p1}, \ref{p2}, \ref{p3}) of the Nice Properties: it
        has many duplicated right-hand sides, it is visually dissimilarity to
        the algebraic laws, and the function, even though it uses pattern
        matching, doesn't really play nice with my custom type; deconstructing
        it is tedious and redundant.         
        
        \begin{figure}
            \begin{center}
                \begin{verbatim}
        let game_of_token token = match token with 
            | BattlePass f      -> ("Fortnite", f)
            | ChugJug f         -> ("Fortnite", f)
            | TomatoTown f      -> ("Fortnite", f)
            | HuntersMark f     -> ("Bloodborne", 2 * f)
            | SawCleaver f      -> ("Bloodborne", 2 * f)
            | MoghLordOfBlood f -> ("Elden Ring", 3 * f)
            | PreatorRykard f   -> ("Elden Ring", 3 * f)
            | _                 -> ("Irrelevant", 0)
                \end{verbatim}
            \end{center}    

        \caption{\tt{game\_of\_token}, with redundant right-hand sides,
        should raise a red flag.} 
        \Description{An implementation of game_of_token using bare patterns.}
        \label{fig:baregot}
        \end{figure}

        I could try to use a couple of helper functions to reduce clutter, and
        write something like Figure~\ref{fig:helpergot}. It looks ok, but I'm
        still hurting for Nice Properties \ref{p2} and \ref{p3}, and the
        additional calls hurt performance. 

        \begin{figure}
            \begin{center}
                \begin{verbatim}
                  let fortnite   f = ... complicated   ... in
                  let bloodborne f = ... complicated'  ... in
                  let eldenring  f = ... complicated'' ... in
                  match token with
                  | BattlePass f -> fortnite f
                  ... and so on ...                
                \end{verbatim}
            \end{center}    
        \caption{\tt{game\_of\_token} with helpers is somewhat better, but I'm
        not satisfied with it.} 
        \Description{An implementation of
        game_of_token using helper functions.}
        \label{fig:helpergot}
        \end{figure}

        Thankfully, an extension once again comes to the rescue.
        \it{Or-patterns} condense multiple patterns which share right-hand
        sides, and I can exploit them to restore the Nice Properties and
        eliminate much uninteresting code in Figure~\ref{fig:orgot}.

    \begin{figure}
    \begin{center}
    \begin{verbatim}
let game_of_token token = match token with 
    | BattlePass f | ChugJug f | TomatoTown f  -> ("Fortnite", f)
    | HuntersMark f | SawCleaver f             -> ("Bloodborne", 2 * f)
    | MoghLordOfBlood f | PreatorRykard f      -> ("Elden Ring", 3 * f)
    | _                                        -> ("Irrelevant", 0)
    \end{verbatim}
    \end{center}    
    \caption{Or-patterns condense \tt{game\_of\_token}
    significantly, and it is easier to read line-by-line.} 
    \Description{An
    implementation of game_of_token using or-patterns.}
    \label{fig:orgot}
    \end{figure}

    In addition to the inherent appeal of brevity, or-patterns serve to
    concentrate complexity at a single juncture and create single points of
    truth. And by eliminating helper functions, like the ones in
    Figure~\ref{fig:helpergot}, or-patterns actually boost performance.
      
    \subsubsection{Wrapping up pattern matching and extensions}
    
    Now, any pattern matching-loving functional programmer like me or (now,
    hopefully!) you knows about the extensions that make pattern matching more
    expressive and how to use them effectively. Earlier, though, you might have
    noticed a problem. Say we face a decision-making problem that obliges us to
    use all of our extensions in unison. When picking a language to do so, we
    are stuck! Indeed, no major functional language has all three of these
    extensions. Remember when we had to switch from OCaml to Haskell to use
    guards, and back to OCaml for or-patterns? The two extensions are mutually
    exclusive in Haskell and OCaml, and also Scala, Erlang/Elixir, Rust, F\#,
    and Agda\cite{haskell, ocaml, scala, erlang, elixir, rust, fsharp, agda}. 


    I find the extension story somewhat unsatisfying. At the very least, I want
    to be able to use pattern matching, with the extensions I want, in a single
    language. Or, I want an alternative that gives me the expressive power of
    pattern matching with these extensions. 

    Last year, I spotted an intriguing alternative to pattern matching:
    \it{equations}, from the Verse Calculus\cite{verse}. Equations present a
    different, yet powerful, way to write code that makes decisions and
    deconstructs data. In the next section, I'll introduce you to equations
    similarly to I how I introduced to you pattern matching. Once you're
    familiar with equations, you'll be ready to compare their strengths and
    weaknesses with those of pattern matching, and judge the compromise I
    propose. 

\subsection{Verse's equations}
    \label{verseoverobservers}

    % \1 Equations 
    Allow me to turn your attention to the equations of the Verse Calculus
    (\VC), a core calculus for the functional logic programming
    language\cite{antoy2010functional, hanus2013functional} Verse\cite{verse}.
    (For the remainder of this paper, I use “\VC” and “Verse” interchangeably.)

    Even if you read the Verse paper, \VC's equations and the paradigms of
    functional logic programming might look unfamiliar to your pattern
    matching-accustomed brains. To help ease you into familiarity with
    equations, I'll ground explanations and examples in how they relate to
    pattern matching. 

    \VC\ uses \it{equations} instead of pattern matching to test for structural
    equality and create bindings. Like pattern matching, equations scrutinize
    and deconstruct data at runtime by testing for structural equality and
    unifying names with values. Unlike pattern matching, \VC's equations can
    unify names on both left- \it{and} right-hand sides. 

    Equations in Verse take the form \it{x = e}, where \it{x} is a name and
    \it{e} is an expression. During runtime, both \it{x} and any unbound names
    in \it{e} are unified with values. Much like pattern matching, unification
    can fail if the runtime attempts to bind incompatible values or structures. 

    Equations offer a form of binding that looks like a single pattern match.
    What about a list of many patterns and right-hand sides, as in a case
    expression? For this, \VC\ has \it{choice} (\choice). The full semantics of
    choice are too complex for me to cover in this paper, but you can get away
    with knowing that choice, when combined with the \tt{one} operator, has a
    very similar semantics to case; that is, “proceed and create bindings if any
    one of these computations succeed.” 

    Let's look at what equations, \tt{one}, and choice look like in real,
    written Verse code. I've written our \tt{area} function in \VC\ extended
    with a \tt{float} type and a multiplication operator \tt{*} in
    Figure~\ref{fig:versearea}. 

    \begin{figure}[]
        \verselst
        \begin{lstlisting}[numbers=none]
$\exists$ area. area = $\lambda$ sh. 
  one {  $\exists$ x. sh   = $\langle$SQUARE, s$\rangle$; s * s
       | $\exists$ w h. sh = $\langle$TRIANGLE, w, h$\rangle$; 
              0.5 * w * h
       | $\exists$ b1 b2 h. sh = $\langle$TRAPEZOID, b1, b2, h$\rangle$; 
              0.5 * b1 * b2 * h}
        \end{lstlisting}
    \caption{\tt{area} in \VC\ uses existentials and equations rather than
    pattern matching.} 
    \Description{An implementation of area in Verse.}
    \label{fig:versearea}
    \end{figure}

    Like in the pattern-matching example in good old \ref{fig:pmarea}, the
    right-hand sides of \tt{area} are \it{guarded} by a “check;” now, the check
    is successful unification in an equation rather than a successful match on a
    pattern. Similarly, \tt{one} with a list of choices represents matching on
    any \it{one} pattern to evaluate a single right-hand side. 

    Why use equations? Let's start with a digestible claim: I prefer \VC's
    equations to observers, and I imagine many functional programmers would as
    well. To support this claim, I appeal to the Nice Properties: 

    \begin{enumerate}
      \item \tt{area} using equations looks like the algebraic laws, with the
      addition of the explicit $\exists$. It's a bit more math-y, but that might
      not be a bad thing: though it doesn't resemble the algebraic laws a
      programmer would write, it likely resembles the equations that a
      mathematician would. 
      \item \tt{area} using equations does not duplicate virtually any code. 
      \item \tt{area} using equations deconstructs user-defined types as easily
      as \ref{fig:pmarea} does with pattern matching. 
      \item \tt{area} using equations has all important internal values named
      very explicitly.
      \item This Property may not hold, because \VC\ on its own is untyped.
      Without a type system, a compiler cannot help me with non-exhaustive
      cases. However, there is no published compiler, type system or no, for
      Verse, and only when one is made available can I make this assertion. For
      this reason, and for the sake of focusing on more important details, I
      choose to proceed in my analysis of equations in \VC\ without considering
      this Property. 
    \end{enumerate}

    If you still believe these Properties to be desirable, you understand why I
    claim programmers prefer equations to observers. Now I'll make a stronger
    claim: equations are \it{at least as good as} pattern matching with popular
    extensions. How can I claim this? By appealing again to the Nice Properties!
    Do you remember how in Section~\ref{pmoverobservers}, I demonstrated how
    pattern matching had to resort to extensions to regain the Properties when
    challenging examples stole them away? In Figure~\ref{fig:verseextfuncs},
    I've implemented those examples in \VC\ (this time extended with strings,
    floats, and \tt{*}) using choice and equations. Take a look for yourself! 

    \begin{figure}[ht] 
        \begin{minipage}[h]{0.54\linewidth}
          \verselst
          \begin{lstlisting}[numbers=none, basicstyle=\tiny, xleftmargin=.2em,
                            showstringspaces=false,
                            frame=single]
$\exists$ exclaimTall. exclaimTall = $\lambda$ sh. 
  one { 
      $\exists$ s. sh = $\langle$Square s$\rangle$; 
      s > 100.0; "Wow! That's a tall square!"
    | $\exists$ w h. sh = $\langle$Triangle, w, h$\rangle$; 
      h > 100.0; "Goodness! Towering triangle!"
    | $\exists$ b1 b2 h. sh = $\langle$Trapezoid, b1, b2, h$\rangle$;
      h > 100.0; "Zounds! Tremendous trapezoid!"
    | "Your shape is small." }
            \end{lstlisting}
                \Description{exclaimTall}
        \subcaption{\tt{exclaimTall} in \VC\ }
            \label{fig:verseexclaimtall} 
        %   \vspace{4ex}
        \end{minipage}%%
        \begin{minipage}[h]{0.5\linewidth}
          \verselst
          \begin{lstlisting}[numbers=none, basicstyle=\tiny, xleftmargin=2em,
                        frame=single]
$\exists$ tripleLookup. tripleLookup = $\lambda$ rho x. 
    one { $\exists$ w. lookup rho x = $\langle$Just w$\rangle$; 
          $\exists$ y. lookup rho w = $\langle$Just y$\rangle$; 
          $\exists$ z. lookup rho y = $\langle$Just z$\rangle$;
          z
        | handleFailure x }
          \end{lstlisting}
                \Description{exclaimTall}
            \subcaption{\tt{tripleLookup} in \VC\ }
            \label{fig:versetriplelookup} 
        \vspace{4ex}
        \end{minipage} 
        \begin{minipage}[h]{\linewidth}
          \verselst
          \begin{lstlisting}[numbers=none, basicstyle=\tiny, xleftmargin=9em, 
                            frame=single, showstringspaces=false]
$\exists$ game_of_token. game_of_token = $\lambda$ token. 
  $\exists$ f. one {
         token = one { $\langle$BattlePass, f$\rangle$ | $\langle$ChugJug, f$\rangle$ | $\langle$TomatoTown, f$\rangle$}; 
           $\langle$"Fortnite", f$\rangle$
       | token = one { $\langle$HuntersMark, f$\rangle$ | $\langle$SawCleaver, f$\rangle$}; 
           $\langle$"Bloodborne", 2 * f$\rangle$
       | token = one { $\langle$MoghLordOfBlood, f$\rangle$ | $\langle$PreatorRykard, f$\rangle$}; 
           $\langle$"Elden Ring", 3 * f$\rangle$
       |   $\langle$"Irrelevant", 0$\rangle$ }
          \end{lstlisting}
                \Description{game_of_token in \VC}
        \subcaption{\tt{game\_of\_token} in \VC\ }
            \label{fig:versegot} 
        \vspace{4ex}
        \end{minipage}%% 
    \caption{Code for the \ref{pmoverobservers} functions with equations looks
    similar, and it doesn't need extensions.}
    \label{fig:verseextfuncs}
      \end{figure}
        
    The code in Figure~\ref{fig:verseextfuncs} has all the Nice Properties
    (again, disregarding the ambiguous 5th.) This is promising for \VC. If it
    rivals pattern matching with popular extensions in desirable properties, and
    \VC\ does everything using only equations and choice, it seems like the
    language is a strong option for writing code! 

    \subsection{\VC\ has a challenging cost model}
    \label{vcbadcost}

    So what's the catch? Programmers everywhere have not thrown up their hands,
    renounced pattern matching, and adopted a puritanical dogma of equations. 
    Sadly, this is not merely a matter of preference. 

    \VC's equations appear to be comparably pleasing to pattern matching in
    their brevity and expressiveness. However, full Verse allows computations
    that are problematic, cost-wise. In \VC, names (logical variables) are
    \it{values}, and can just as easily be the result of evaluating an
    expression as an integer or tuple. To bind these names, \VC, like other
    functional logic languages, relies on \it{unification} of its logical
    variables and \it{search} at runtime to meet a set of program
    constraints\cite{antoy2010functional, hanus2013functional}. Combining
    unifying logical variables with search at runtime classically requires
    backtracking, which can lead to exponential runtime
    cost\cite{hanus2013functional, wadler1985replace, clark1982introduction}. 

    % Todo: An example would be nice. Backtracking/evaluating backwards, then
    % done. 

    Pattern matching, by contrast, has a desirable cost model.
    Maranget\cite{maranget} showed that pattern matching can be compiled to a
    \it{decision tree}, a data structure that enforces linear runtime
    performance by guaranteeing no part of the scrutinee will be examined more
    than once. A decision tree, however, forbids backtracking by nature: once
    the program makes a decision based on the form of a value, it can't re-test
    it later with new information. 
    

\end{document}