\documentclass[manuscript,screen,review, 12pt, nonacm]{acmart}
\let\Bbbk\relax % Fix for amssymb clash 
\usepackage{vmlmacros}
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}
\usepackage{outlines}
\setlength{\headheight}{14.0pt}
\setlength{\footskip}{13.3pt}
\title{An Alternative to Pattern Matching, Inspired by Verse}

\author{Roger Burtonpatel}
\email{roger.burtonpatel@tufts.edu}
\affiliation{%
\institution{Tufts University}
\streetaddress{419 Boston Ave}
  \city{Medford}
  \state{Massachusetts}
  \country{USA}
  \postcode{02155}
  }
\bibliographystyle{ACM-Reference-Format}
\usepackage[utf8]{inputenc}

\begin{document}

\section{Acknowledgements}

This thesis would not have been possible without the infinitely generous time
and support of my advisors, Norman Ramsey and Milod Kazerounian. Norman's
offhand comment of “I~wonder if Verse's equations subsume pattern matching” was
the entire basis of this work, and his generosity in agreeing to advise a full
thesis to answer his question will always be profoundly appreciated. During the
academic year, Norman provided me with materials on both the technical story and
on how to write about it well. He gave me regular feedback and has helped
improve my research skills, my technical writing, and my understanding of
programming languages in general tremendously. I~especially appreciate how he
has guided me in-person at the end of my undergraduate when his book~\citep{bpc}
got me started down the path of PL at the beginning of it. Finally, Norman is
also fantastically fun to pair program with. 

From the beginning, Milod provided me with encouraging mentorship that kept
me enthusiastic and determined to complete the project. He was exceptionally
patient as I~gave him whirlwind tour after whirlwind tour of the changing
codebase and problems, and kept me grounded in the problems at hand. He 
sent me helpful examples of his research to aid me in my proofs, and gave me
some particularly encouraging words towards the end of the project that I~
will not soon forget. 

My undergraduate advisor, Mark Sheldon, has always been both supportive and
kind. I~have enjoyed many long talks in his office, and I~am deeply grateful
that he is on my committee. 

Alva Couch was the advisor of my original thesis idea, which was to compile
programming languages with music. Ultimately, we decided that I~should pursue
this project instead, and I~am grateful to him for his mentorship in that moment
and onwards. 

My family--- my mother, Jennifer Burton, my father, Aniruddh Patel, and my
sister, Lilia Burtonpatel, have all given me support, encouragement, and
(arguably most importantly) food. My gratitude for them is immeasurable. 

My gratitude towards my friends is also without limit. In particular, and in
no order, Liam Strand, Annika Tanner, Max Stein, Cecelia Crumlish, and
Charlie Bohnsack all showed specific interest in the work and encouraged me
as I~moved forward. Rachael Clawson was subjected to much rubber-ducking,
and endured valiantly. Aliénor Rice and Marie Kazibwe were as steadfast
thesis buddies as I~could ever hope for. Jasper Geer, my PL partner in
crime, was always one of my favorite people to talk to about my thesis. His
pursuits in research inspire mine. 

Skylar Gilfeather, my unbelievable friend. Yours is support that goes
beyond words; care, food, silent and spoken friendship, late night rides to
anywhere, laughs and tears are some that can try to capture it. I~am so, so
grateful for how ceaselessly you've encouraged me on this journey. Every one
of your friends is lucky to have you in their life, and I~am blessed that 
you are such a core part of mine. 

Anna Quirós, I~am writing these words as you sleep behind me. Your support
and love have been immeasurable. I~will always be grateful to you for this
year \it{increíble}. I~could not have smiled through it all without you. 

Thank you all. 

\section{References}
\bibliography{sources}

\renewcommand\thesection{\Alph{section}}
\setcounter{section}{0}
\section{Proofs}
\begin{outline}
\1 \bf{Proof: Translation from \VMinus to \D\ preserves semantics }
\begin{proof}
  In progress. 
\end{proof}
\1 \bf{Proof: Translation from \PPlus\ to \VMinus preserves semantics }
\begin{proof}
  In progress. 
\end{proof}
% \1 \bf{Proof: Translation from \VMinus to Verse preserves semantics}
\end{outline}

\section{Discussion}

Both backtracking and multiple results often manifest within \VC's choice
operator, which tempt its removal from \VMinus altogether. However, I~am drawn
to harnessing the expressive potential of this operator, particularly when
paired with \VC's 'one' keyword. When employed with choice as a condition, 'one'
elegantly signifies 'proceed if any branch of the choice succeeds. 

To this end, in \VMinus, choice is permitted with several modifications:

\begin{enumerate}
\item Choice may only appear as a condition or 'guard', not as a result
or the right-hand side of a binding.
\item If any branch of the choice succeeds, the choice succeeds,
producing any bindings found in that branch. The program examines the
branches in a left-to-right order.
\item The existential $\exists$ may not appear under choice.
\end{enumerate}

I~introduce one more crucial modification to the \VC runtime: a name in
\VMinus is an \it{expression} rather than a \it{value}. This alteration,
coupled with my adjustments to choice, eradicates backtracking. Our
rationale behind this is straightforward: if an expression returns a
name, and subsequently, the program imposes a new constraint on that
name, it may necessitate the reevaluation of the earlier expression—- a
scenario I~strive to avoid. 

\subsubsection{\VMinus and \PPlus, side by side}
I~compare \VMinus with \PPlus as an exercise in comparing equations with
pattern matching. They certainly look similar, which hints that \VMinus
might be as expressive as pattern matching with the three extensions. 
Proving this claim is the topic of Section~\ref{pplustovminus}. 

When might a programmer prefer \VMinus over \PPlus, or vice versa? After
programming in both, I~have come up with three empirical observations: 

\begin{enumerate}
  \item \it{The scrutinee:} When there is no obvious single scrutinee, \VMinus is more succinct.
  When there is a scrutinee, \PPlus is more succinct. 
  \item \it{Binding and decision-making:} Binding and decision-making are
  joined in a single construct in \VMinus: \tt{=}. \PPlus needs different
  kinds of syntax, like \tt{<-}, to express different kinds of binding. And
  a programmer wanting for \tt{let} in \PPlus will never feel this need in
  \VMinus: \tt{=} subsumes that, too. 
  \item \it{Names:} In \VMinus, names are explicitly introduced. In
  practice, this helps prevent a common mistake in pattern matching in which
  a programmer attempts to match a value $v$ on an in-scope name $x$,
  expecting the match to succeed iff $x$ evaluates to $v$ at runtime, only
  to see the match always succeed. 
\end{enumerate}     

% \rab{How do you feel about examples for each of these points here? This
% section is very long, but they help drive these points home.}


% An example in \VC illustrates this precise
% scenario: 

% FIRST PAPER EXAMPLE 

% IMPORTANTLY, \VMinus REALLY ISN'T A FUNCTIONAL LOGIC PROGRAMMING LANGUAGE ANYMORE. 

%  Doesn't backtrack

%  No multiple results 

%  Doesn't evaluate functions backwards, have top-level patterns like verse, the list goes on 

% If you're well-versed in functional logic programming, may perceive that
% imposing such restrictions on choice and names effectively strips away
% much of Verse's essence as a functional logic programming language. With
% these constraints enforced, there can be no backtracking, multiple
% results, backward function evaluation, or top-level patterns, among
% other classic functional logic programming features. But do not fear:
% our intent is not to recklessly strip Verse of its meticulously crafted
% core tenets. Instead, our aim is to extract a select few—namely, its
% equations, existentials, and nondeterministic evaluation order—and
% juxtapose them with pattern matching.    

Indeed, having stripped out the functional logic programming elements of
\VC, only the decision-making bits are left over. To wrap these, I~add a
classic decision-making construct: guarded commands~\citep{dijkstra} The result is \VMinus. \VMinus's concrete syntax is
defined in Figure~\ref{fig:vmsyntax}. 

\section{Addressing how \PPlus handles unusual pattern combinations}
\label{ppweird}
    \PPlus admits of strange-looking patterns: consider \tt{Cons (when true)
    zs}. But these should not be alarming, because such syntactic forms reduce
    to normal forms by (direct) application of algebraic laws: 

    \begin{align}
      K (\whenexpr) \;p2\; \dots &=== K \;\_ \;p2\; \dots, \whenexpr \\
      K (\whenexpr, \;p2\;) \;p3\; \dots  &=== K \;p2\; \;p3\; \dots, \whenexpr \\
      K (p1\;, \whenexpr) \;p3\; \dots  &=== K \;p1\; \;p3\; \dots, \whenexpr \\
      K (\whenexpr \pbar p2\;) \;p3\; \dots &=== (K \;\_ \;p3\; \dots, \whenexpr) \pbar (K \;p2\; \;p3\; \dots) \\
      K (p1 \pbar \whenexpr) \;p3\; \dots &=== (K \;p2\; \;p3\; \dots) \pbar (K \;\_ \;p3\; \dots, \whenexpr)  \\
      \whenexpr \leftarrow e &=== \;\_ <- e, \whenexpr
    \end{align}   
    
    Repeatedly applying these laws until the program reaches a fixed point
    normalizes placements of \it{when}. Laws (2) and (3) work because \PPlus
    has no side effects and the laws assume all names are unique (the compiler
    takes care of this), so changing the order in which patterns match has no
    effect on semantics.         

\section{Is \VMinus\ a true subset of \VC?}
\VMinus\ certainly appears to relate to \VC\ semantically. Translating
\it{if-fi} and choice in \VMinus to \bf{one} and choice in \VC\ is likely a
sufficient embedding. Formalizing this translation and, more importantly,
proving that our semantics of \VMinus are consistent with Augutsson et al.'s
\VC\ is an excellent goal for future work. 

\newcommand\dbar{▯}
\section{Formal Definitions of all languages}
\label{languagedefs}

\begin{table}[ht]
  \centering
  \small
  \begin{tabular}{l l l}
      \textbf{Syntactic Forms} & \textbf{Cases} & \textbf{Unique to} \\
      \hline
      $P$ : Programs        & $\bracketed{d}$        & \\
      $d$ : Definitions     & \valabstract           & \\
      $\expr$ : Expressions & $\v$                   & \\
                            & $x, y, z$             & \\
                            & $K\bracketed{\expr}$  & \\
                            & $\lambda x.\; \expr$  & \\
                            & $\expr[1]\; \expr[2]$ & \\
                            & \caseabstract         & \PPlus \\
                            & \iffiabstract         & \VMinus \\
                            & $\dt$                 & \D \\
      \\
      $\v$ : Values          & $K\bracketed{\v}$     & \\
                            & $\lambda x.\; \expr$  & \\
      \\
      $p$ : Patterns         & $x$                  &  \\
                            & $\_$                  &  \\
                            & $K\; \bracketed{p}$   &  \\
                            & $\whenexpr$           & {\smash{$\left . \vrule height 7ex depth 7ex width 0pt \right\}\; \PPlus$}} \\
                            & $\pcommap$            & \\
                            & $\parrowe$            & \\
                            & $\porp$               & \\
      \\
      $\G$ : Guarded Expressions & \gexpabstract      & \\
      $\g$ : \textsf{Guards}  & $x = \expr$           & \raisebox{-0.3em}{\smash{$\left . \vrule height 5ex depth 5ex width 0pt \right\}\; \VMinus$}} \\
                            & $\expr$               & \\
                            & \choiceabstract       & \\
                            % need raisebox here 0.5\baselineskip from "one below middle"
      \\                            
      $\dt$ : Decision Trees & $\decisiontreeabstract$ & \\
                             & $\expr$                    & \\
                             & \treetryletabstract        & {\smash{$\left . \vrule height 8ex depth 6ex width 0pt \right\}\; \D$}} \\
                             & \treeifabstract           &  \\
                             % & \treeextractabstract       & \\
                             & \treefailabstract          & \\
      \\
  \end{tabular}
  \caption{Abstract Syntax of all languages, 
           with delineations in column \bf{Unique To}}
  \label{fig:unilang}
\end{table}
\ppsemantics
\vmsemantics

% \begin{figure}
% \begin{center}
%     \begin{bnf}
%     $P$ : \textsf{Programs} ::=
%     $\bracketed{d}$ : definition
%     ;;
%     $d$ : \textsf{Definitions} ::=
%     | $\tt{val}\; x\; \tt{=}\; \expr$ : bind name to expression
%     ;;
%     $\expr$ : \textsf{Expressions} ::=
%     | $v$ : literal values 
%     | $x, y, z$ : names
%     | $\tt{if}\; [\; G\; \bracketed{\tt{[]}\; G}\;]\; \tt{fi}$ : if-fi 
%     | $K \bracketed{\expr}$ : value constructor application 
%     | $\expr[1]\; \expr[2]$ : function application 
%     | $\ttbackslash x\tt{.}\; \expr$ : lambda declaration 
%     ;;
%     $G$ : \textsf{Guarded Expressions} ::=  
%     $[\tt{E }{\bracketed{x}}{\tt{.}}] \bracketed{g} \;\ttrightarrow\; \expr$ : names, guards, and body
%     ;;
%     $g$ : \textsf{Guards} ::=  
%     | $\expr$ : intermediate expression 
%     | $x \;\tt{=}\; \expr$ : equation 
%     | $ g \bracketed{\tt{;} g} \; \pbar \; g \bracketed{\tt{;} g}$ : choice 
%     ;;
%     $\v$ : Values ::= $K\bracketed{\v}$ : value constructor application 
%     | $\ttbackslash x\tt{.}\; \expr$ : lambda value
%     \end{bnf}
% \end{center}
% \Description{The concrete syntax of PPlus.}
%     \caption{\VMinus: Concrete syntax}
%     \label{fig:vmsyntax}
%     \end{figure}

% \begin{figure}
%     \begin{center}
%     \begin{bnf}
%     $P$ : \textsf{Programs} ::=
%     $\bracketed{d}$ : definition
%     ;;
%     $d$ : \textsf{Definitions} ::=
%     | $\tt{val}\; x\; \tt{=}\; \expr$ : bind name to expression
%     ;;
%     $\expr$ : Expressions ::= 
%     | $v$ : literal values 
%     | $x, y, z$ : names
%     | $K\bracketed{\expr}$ : value constructor application 
%     | $\ttbackslash x\tt{.}\; \expr$ : lambda declaration  
%     | $\expr[1]\; \expr[2]$ : function application 
%     | $\tt{case}\; \expr\; \bracketed{p\; \ttrightarrow\; \expr}$ : case expression 
%     | \ttbraced{$\expr$}
%     ;;
%     $p$ : \textsf{Patterns} ::= $p_{1}\pbar p_{2}$ : or-pattern
%     | $p \tt{,} p'$ : pattern guard 
%     | $p\; \tt{<-}\; \expr$ : pattern from explicit expression  
%     | $x$ : name 
%     | $\tt{\_}$ : wildcard 
%     | $K\; \bracketed{p}$ : value constructor application 
%     | $\tt{when}\; \expr$
%     | \ttbraced{$p$}
%     ;;
%     $\v$ : Values ::= $K\bracketed{\v}$ : value constructor application 
%     | $\ttbackslash x\tt{.}\; \expr$ : lambda value 
%     ;;
%     $K$ : \textsf{Value Constructors} ::=
%     | $\tt{true}\; \vert\; \tt{false}$ : booleans
%     | $\tt{A-Z}x$ : name beginning with capital letter
%     % | $[\tt{-}\vert\tt{+}](\tt{0}-\tt{9})+$ : signed integer literal 
%     \end{bnf}
%     \end{center}
%     \Description{The concrete syntax of PPlus.}
%     \caption{\PPlus: Concrete syntax}
%     \end{figure}

\end{document}
