\documentclass[manuscript,screen, 12pt, nonacm]{acmart}
\title{Response Letter}
\let\Bbbk\relax % Fix for amssymb clash 
\usepackage{vmlmacros}

\begin{document}
Dear Norman, 

Below is my response to your suggested revisions for my thesis, \textit{An
Alternative to Pattern Matching, Inspired by Verse}. Thank you for your
suggestions on this document. 

\it{Page 1}

I have made suggested revisions to the abstract, fixed the flow and citation
order issues in the introduction, and added a table of contents. 

\it{Page 2}

I have made the suggested cuts, edits, and replacements. I have removed the
unsupported claim of subsumption. 

\it{Page 3}

I have fixed the \tt{shape} datatype to be called \tt{standard\_shape}, and
given a suitable explanation. I have fixed the trapezoid area formula (!) here
and everywhere. I have rewritten the garden-path sentence. I have made the
suggested cuts, edits, and replacements.

\it{Page 4}

I have given the Nice Properties names. I have made the suggested cuts, edits,
and replacements. I have changed the order to present first, discuss after. 

\it{Page 5}

I have revised the Nice Properties to be supported, especially the
Exhaustiveness property, and given the figures names. I have fixed the odd
spacing around my em dashes. I have made the suggested cuts, edits, and
replacements.

\it{Page 6}

I have changed future tense to present tense on this and future pages where
applicable. I have made the suggested cuts, edits, and replacements.

\it{Page 7}

I have fixed the \tt{exclaimBigArea} typo. I like your suggestion of “sizeable
square,” and have implemented it. I have made the suggested cuts, edits, and
replacements. 

\it{Page 8}

I have propagated edits from above to this page. I have kept the last paragraph
as-is, and am open to more specific suggestion on how to edit it. I have made
the suggested cuts, edits, and replacements. 

\it{Page 9}

I have made the “too mysterious” paragraph more concrete, and point to the
example in the next section. I have made the suggested cuts, edits, and
replacements. 

\it{Page 10}

I have propagated edits from above to this page. I have made the suggested cuts,
edits, and replacements. 


\it{Page 11}

I have chosen not to drop the “could drop” paragraph, because I find it
informative. I have fleshed out what a 'token' means. I have made the suggested
cuts, edits, and replacements. 

\it{Page 12}

I have made the suggested cuts, edits, and replacements. 

\it{Page 13}

I have made the suggested cut, and added in its place a Nice Property citation.


\it{Page 14}

I have fixed the citations. I have split Section 2 into two sections, as you
suggested: PM and \VC. I have cut the paragraph labeled “Discussion, it at all.”
I have made the suggested cuts, edits, and replacements. 

\it{Page 15}

I have added a citation for Robinson's Unification Algorithm. I have added a
note about the full equation form, $\expr[1] = \expr[2]$. I have fixed the
garden-path sentence on observers. I have displayed the~\VC examples alongside
the algebraic laws for the reader to judge. I have a walkthrough of each keyword
and symbol used in the examples. I explain what I meant by “mathematical notation.”
I have made the suggested cuts, edits, and replacements. 

\it{Page 16}

I have fixed the “Weak” Property. I have added text that more extensively
discusses Exhaustiveness in \VC and in Verse. I have fixed the garden-path
sentence on observers. I have added ties here and everywhere.  I have replaced
“strong” with “intriguing.” I have made the suggested cuts, edits, and
replacements. 

\it{Page 17}

I have removed~\PPlus from the thesis, and now I only discuss~\VMinus and~D. I
have made the suggested cuts, edits, and replacements. 

\it{Page 18}

I have cascaded {\PPlus}'s removal to this and future pages. I have moved the \U
table up significantly, so it does not need a page number. I have also added
hyperlinks to all of my figures and tables, so they should not need page
numbers. I have made the suggested cuts, edits, and replacements. 

\it{Page 19-25}

I have added your suggestion on Peano numbers. After that, everything from here
to Page 25 has been completely removed. 

\it{Page 26} 

I have removed the table entirely. I have made the suggested cuts, edits, and
replacements. 

\it{Page 27} 

I have removed the first Lemma and “Proof” pair, and the second. To address your
long note, I further compare \VMinus and \VC in the Discussion. I have merged
the concrete syntax of \VMinus with the abstract to create a single point of
truth, as per our discussion. The syntax is now unified everywhere. I now use
Dijkstra's definitive \dbar. I have made the suggested cuts, edits, and
replacements. 

\it{Page 28} 

\bf{I have not found a way to put the examples side-by-side in a way that is both
satisfying and legible to me, and would like further guidance.} Specifically, I
want to put all three of pattern matching, \VC, and \VMinus together, and would
like input on how best to do that. I have fixed the copypasta. I have made the
suggested replacement of “a” for “the.”

\it{Page 29} 

I have fixed the metavariables table by adding “An” or “A” to each row. I have
added \Gs~and \gs to the table. I have fixed the font and made it clear that
\fail is a syntactic form. I have eliminated the context~\script{T}\; from
\VMinus. I have not moved the judgement rule that says “Not here”, as I am not
sure why it fits exactly there. \bf{I would like further input on this.} I have made
the other suggested cuts, edits, and replacements. 

\it{Page 30} 

Linking to the previous note, I do not understand from where your arrow with
“Judgement Form” is originating. If it is originating from “determined,” I do
not understand why. I have made the other suggested cuts, edits, and
replacements. I have moved refinement earlier, and given it an explanation. 

\it{Pages 31-34} 

I have added a short narrative to accompany the rules you pointed out in
\VMinus. I have fixed the typos and made the suggested replacements. I have
added closures to lambda. \bf{If you are still willing to send the mathpar example
to box the rules, I would like it.}

\it{Page 34}

I have moved \D to the next section. I have made all the suggested cuts, edits,
and replacements on this page. 

\it{Page 35}

I have changed the order to introduce \D fully before the compiler. I have made
the claim more reasonable: I say, “To demonstrate that \VMinus has a
similarly-desirable cost model to pattern matching\dots” I have moved the marked
paragraph to the conclusion. I have moved the paragraph where I discuss
differences between my and Maranget's algorithm here. I have made all the
suggested cuts, edits, and replacements. 

\it{Page 36}

I explain why I use Maranget's tree instead of \D (this is in the paper): 

“I show Maranget's example for two reasons: First, the example serves to bolster
your understanding of decision trees with a classic, well-established model.
Second, since I currently have no visualization generator for~\D (\D~can
currently only be visualized as plain text), I use Maranget's example to have a
reasonable visual representation of decision trees.”

I have gotten rid of the \it{extract} node. 

I have reordered the sections to make sense. 

I have made all the suggested cuts, edits, and replacements. 

\it{Page 37}

I have merged the concrete syntax of \D with the abstract to create a single
point of truth, as per our discussion.  The syntax is now unified everywhere.
\it{extract} is gone, and I have taken inspiration from \it{Exceptional Syntax}
in redesigning \it{try-let} to \it{let-unless}. I have rewritten the compilation
section. 

\it{Page 38-40}

To signpost more: I have rewritten the compilation section as per your
suggestion to use nondeterministic application of rules. I give a short
description of the rules, as well, in addition to the full, uninterrupted rules
at the end of the section. For the proof, I give the idea with the theorem. 

\it{Page 41-43}

I have removed the claim of subsumption, and removed this section (5) entirely. 

\it{Page 43}

I have fleshed out what I actually did with the implementations. I have merged
related and future work, since I found them very closely intertwined in this
thesis and believe they work well together here. I have written more on
alternatives to big-step semantics. I have included Maranget's translation
algorithm. I have added the tie. I have made all the suggested cuts, edits, and
replacements. 

\it{Page 44}

I have removed the “Using \VMinus\dots” section. I have made all the suggested
cuts, edits, and replacements. 

\it{Page 45}

I have adjusted the conclusion to use the right verb: I \it{compared} \VMinus
with pattern matching and \VC to demonstrate my claim. I have removed the claim
that “I~have also shown that translating from pattern matching to equations
preserves the desirable properties.” I have removed the last sentence of the
conclusion, which made me sad. I have fixed “undergraduate” to be “undergrad”

\it{Page 46}

I have added “he.”

\it{Page 47}

I have fixed the duplicate \bf{References} title, and fixed the individual
reference items as per our discussion. 

\it{Page 48}

I have removed the incomplete \bf{Proofs} section. I have fixed the verbs in the
Discussion. I have edited the “\VMinus vs. \VC” section to justify my
discussion. I would like to include more detailed analysis in the future. 

\it{Page 49-50}

Having removed \PPlus, I have removed the sections that mention it. 

\it{Page 50}

I have given a reason to care about the subset. I have moved Appendix E's
contents to the main text in their own section, and I also have the Appendix for
reference. Commentary appears more in the body of the paper in the newest draft,
though I am happy to add more commentary here. 

\it{Page 51}

I have cleaned up \U, including color-coding, and moved it to the front.

\it{Page 52}

I have fixed lambda. I would like your examples for nicely condensing rules. 

\it{Page 52-54}

I have removed all the rules for \PPlus. 

\it{Page 54}

There is no more context, so rule \textsc{Move-Guard-To-Ctx} is gone. 

\it{Page 55}

I have replaced \textsc{Assoc} with \textsc{Commut}. I have switched the order
on the \textsc{(Guard-Names-Bot-Succ)} rules, which makes more sense. 

\it{Page 56}

I have fixed the typos on this page, changed $\square$ to \dbar, and \dots to
\Gs. 

\it{Page 57}

Ditto on typos, changes, and lambda fixes. 

\it{Page 58}

Only page in document with no edits. 


With sincere thanks, 
Roger

\end{document}